---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# phyr

The goal of phyr is to collect and update (with c++ for core parts) functions that:

- calculate alpha phylogenetic diversity (`psv`, `psr`, `pse`, etc.) and beta phylogenetic diversity (`pcd`) from the picante package
- fitting phylogenetic logistic regressions (`binaryPGLMM`) from the ape package
- fitting phylogenetic generalized linear mixed models (`communityPGLMM`) from the pez package
- and more.

These functions share some similarities and it makes more sense to put them in one package to reduce redundancy in codes and to facilitate updates.

# To do

- Import `psv` family of functions, change the default pruning setting of the phylogeny since this sometimes can lead to different results from not pruning.
- Import `binaryPGLMM`
- Import `communityPGLMM`

# Imported

- `pcd` from the picante package
- `psv` from the picante package

```{r, message=FALSE}
library(phyr)
# pcd is about 20 times faster
microbenchmark::microbenchmark(phyr::pcd(comm = comm_a, tree = phylotree, reps = 1000, verbose = F),
                               picante::pcd(comm = comm_a, tree = phylotree, reps = 1000),
                               times = 30)
# psv, the example data is too small to compare
microbenchmark::microbenchmark(phyr::psv(comm_a, phylotree),
                               picante::psv(comm_a, phylotree))
```

