---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# phyr

The goal of phyr is to collect and update (with c++ for core parts) functions that:

- calculate alpha phylogenetic diversity (`psv`, `psr`, `pse`, etc.) and beta phylogenetic diversity (`pcd`) from the picante package
- fitting phylogenetic logistic regressions (`binaryPGLMM`) from the ape package
- fitting phylogenetic generalized linear mixed models (`communityPGLMM`) from the pez package
- and more.

These functions share some similarities and it makes more sense to put them in one package to reduce redundancy in codes and to facilitate updates.

# Installation

To install this package:

```{r eval = F}
devtools::install_github("daijiang/phyr")
# or install the binary version
install.packages("https://raw.githubusercontent.com/daijiang/phyr/master/phyr_0.1.2.tgz", repos = NULL)
```


# To do

- update `psv` family of functions

# Imported

- `pcd` from the picante package; changed the default pruning setting of the phylogeny since this sometimes can lead to different results from not pruning.
- `psv` from the picante package
- `communityPGLMM` from the pez package
- `binaryPGLMM` from the ape package

```{r, message=FALSE}
library(phyr)
# pcd is about 20 times faster
microbenchmark::microbenchmark(phyr::pcd(comm = comm_a, tree = phylotree, reps = 1000, verbose = F),
                               picante::pcd(comm = comm_a, tree = phylotree, reps = 1000),
                               times = 30)
# psv, the example data is too small to compare
microbenchmark::microbenchmark(phyr::psv(comm_a, phylotree),
                               picante::psv(comm_a, phylotree))
```

`communityPGLMM` now can use similar syntax as `lme4::lmer` to specify random terms: add `__` (two underscores) at the end of grouping variable (`sp`) to specify both phylogenetic and non-phylogenetic random terms; use `(1|sp@site)` to specify nested term. Note: `(1|sp@site)` and `(1|site@sp)` have different orders.

```{r eval = T}
library(dplyr)
comm = comm_a
comm$site = row.names(comm)
dat = tidyr::gather(comm, key = "sp", value = "freq", -site) %>% 
  left_join(envi, by = "site") %>% 
  left_join(traits, by = "sp")
dat$pa = as.numeric(dat$freq > 0)
head(dat)
test1 = phyr::communityPGLMM(freq ~ 1 + shade + (1|sp__) + (1|site) + (1|sp@site), 
                             data = dat, family = "gaussian", tree = phylotree, REML = F)
test1
test3 = phyr::communityPGLMM(pa ~ 1 + shade + (1|sp__) + (1|site) + (1|sp@site), 
                             data = dat, family = "binomial", tree = phylotree, REML = F)
test3
```

To compare the cpp version and R version, and the version from the `pez` package.

```{r, warning=FALSE}
# data prep for pez::communityPGLMM, not necessary for phyr::communityPGLMM
dat = arrange(dat, site, sp)
nspp = n_distinct(dat$sp)
nsite = n_distinct(dat$site)

dat$site = as.factor(dat$site)
dat$sp = as.factor(dat$sp)

tree = ape::drop.tip(phylotree, setdiff(phylotree$tip.label, unique(dat$sp)))
Vphy <- ape::vcv(tree)
Vphy <- Vphy/max(Vphy)
Vphy <- Vphy/exp(determinant(Vphy)$modulus[1]/nspp)
Vphy = Vphy[levels(dat$sp), levels(dat$sp)]

# prepare random effects
re.site <- list(1, site = dat$site, covar = diag(nsite))
re.sp <- list(1, sp = dat$sp, covar = diag(nspp))
re.sp.phy <- list(1, sp = dat$sp, covar = Vphy)
# sp is nested within site
re.nested.phy <- list(1, sp = dat$sp, covar = Vphy, site = dat$site)
re.nested.rep <- list(1, sp = dat$sp, covar = solve(Vphy), site = dat$site) # equal to sp@site
# can be named 
re = list(re.sp = re.sp, re.sp.phy = re.sp.phy, re.nested.phy = re.nested.phy, re.site = re.site)

# about 4-10 times faster for a small dataset
microbenchmark::microbenchmark(
  phyr::communityPGLMM(freq ~ 1 + shade + (1|sp__) + (1|site) + (1|sp@site), 
                       dat, tree = phylotree, REML = F, cpp = T, optimizer = "bobyqa"),
  phyr::communityPGLMM(freq ~ 1 + shade + (1|sp__) + (1|site) + (1|sp@site), 
                       dat, tree = phylotree, REML = F, cpp = T, optimizer = "Nelder-Mead"),
  phyr::communityPGLMM(freq ~ 1 + shade + (1|sp__) + (1|site) + (1|sp@site), 
                       dat, tree = phylotree, REML = F, cpp = F, optimizer = "Nelder-Mead"),
  pez::communityPGLMM(freq ~ 1 + shade, data = dat, sp = dat$sp, site = dat$site, 
                      random.effects = re, REML = F),
  times = 5
)

# about 6 times faster for a small dataset
microbenchmark::microbenchmark(
  phyr::communityPGLMM(pa ~ 1 + shade + (1|sp__) + (1|site) + (1|sp@site), dat, 
                       family = "binomial", tree = phylotree, REML = F, cpp = T, 
                       optimizer = "bobyqa"),
    phyr::communityPGLMM(pa ~ 1 + shade + (1|sp__) + (1|site) + (1|sp@site), dat, 
                       family = "binomial", tree = phylotree, REML = F, cpp = T, 
                       optimizer = "Nelder-Mead"),
  phyr::communityPGLMM(pa ~ 1 + shade + (1|sp__) + (1|site) + (1|sp@site), dat, 
                       family = "binomial", tree = phylotree, REML = F, cpp = F, 
                       optimizer = "Nelder-Mead"),
  pez::communityPGLMM(pa ~ 1 + shade, data = dat, family = "binomial", sp = dat$sp, 
                      site = dat$site, random.effects = re, REML = F),
  times = 5
)
```


