---
title: "Profvis of communityPGLMM"
author: "Daijiang Li"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Profvis of communityPGLMM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  eval = F
)
```

In this file, we show the memory and time used for `phyr::communityPGLMM()`. It seems that for Gaussian models, use `bobyqa` as optimizer can save time and memory. But for Binomial models, the optimizer does not affect too much. Thus, may use `Nelder-Mead` for its robustness.

Also, note the hugh differences in memory usage between `cpp = TRUE` and `cpp = FALSE`.

First, let's prepare some toy data.

```{r data_prep}
library(profvis)
library(dplyr, quietly = T)
library(phyr)
comm = comm_a
comm$site = row.names(comm)
dat = tidyr::gather(comm, key = "sp", value = "freq", -site) %>% 
  left_join(envi, by = "site") %>% 
  left_join(traits, by = "sp")
dat$pa = as.numeric(dat$freq > 0)
```

Then, we start the profvis.

# Gaussian models

### `bobyqa` with R

```{r prof1}
profvis({
  phyr::communityPGLMM(freq ~ 1 + shade + (1|sp__) + (1|site) + (1|sp@site), dat, 
                       tree = phylotree, REML = T, cpp = F, optimizer = "bobyqa")
})
```

### `bobyqa` with Rcpp

```{r prof2}
profvis({
  phyr::communityPGLMM(freq ~ 1 + shade + (1|sp__) + (1|site) + (1|sp@site), dat, 
                       tree = phylotree, REML = T, cpp = T, optimizer = "bobyqa")
})
```

### `Nelder-Mead` with R

```{r prof3}
profvis({
  phyr::communityPGLMM(freq ~ 1 + shade + (1|sp__) + (1|site) + (1|sp@site), dat, 
                       tree = phylotree, REML = T, cpp = F, optimizer = "Nelder-Mead")
})
```

### `Nelder-Mead` with Rcpp

```{r prof4}
profvis({
  phyr::communityPGLMM(freq ~ 1 + shade + (1|sp__) + (1|site) + (1|sp@site), dat, 
                       tree = phylotree, REML = T, cpp = T, optimizer = "Nelder-Mead")
})
```

# Binomial models

### `bobyqa` with R

```{r prof5}
profvis({
  phyr::communityPGLMM(pa ~ 1 + shade + (1|sp__) + (1|site) + (1|sp@site), dat, 
                       family = "binomial", tree = phylotree, REML = T, cpp = F, 
                       optimizer = "bobyqa")
})
```

### `bobyqa` with Rcpp

```{r prof6}
profvis({
  phyr::communityPGLMM(pa ~ 1 + shade + (1|sp__) + (1|site) + (1|sp@site), dat, 
                       family = "binomial", tree = phylotree, REML = T, cpp = T, 
                       optimizer = "bobyqa")
})
```

### `Nelder-Mead` with R

```{r prof7}
profvis({
  phyr::communityPGLMM(pa ~ 1 + shade + (1|sp__) + (1|site) + (1|sp@site), dat, 
                       family = "binomial", tree = phylotree, REML = T, cpp = F, 
                       optimizer = "Nelder-Mead")
})
```

### `Nelder-Mead` with Rcpp

```{r prof8}
profvis({
  phyr::communityPGLMM(pa ~ 1 + shade + (1|sp__) + (1|site) + (1|sp@site), dat, 
                       family = "binomial", tree = phylotree, REML = T, cpp = T, 
                       optimizer = "Nelder-Mead")
})
```
